### 2010-02-23 trunk/deliverables/D2_2_Ontology/fast_d2_2_2_ontology.tex
review-ready version ready
git-svn-id: https://svn.forge.morfeo-project.org/fast-fp7project@589 d8b0c59b-6544-0410-94d9-83a5658a058e

@@

This deliverable is structured as follows: as part of the introduction, we will present the methodology we chose to adopt for the development of the FAST gadget ontology, by discussing  each of its proposed stages in detail. Following, we will conclude the introduction by elaborating more on some of the general design decisions we are following. After briefly addressing the topic of related work in Sect.~\ref{sec:related_work} --- which is mainly covered in a different deliverable from this work package, \cite{urmetzer2010fast_state_of_the_art} --- the structure of the remainder of this document follows directly from our adopted methodology: we will perform a domain analysis in Sect.~\ref{sec:domain_analysis}, covering the stages of specification, conceptualisation and integration of existing vocabularies. This is then followed by the implementation step in Sect.~\ref{sec:ontology}, which covers the ontology in its final form (at this moment of the project). Finishing the deliverable, we present our conclusions and outlook on years two and three of the project in Sect.~\ref{sec:conclusions}.

@@

@@

We are not aware of any other project which has already developed or planned to develop an ontology of the gadget domain. However, there is a range of related material from which we take inspiration regarding the description of the gadget domain. This material comes from a number of different sources and directions. It includes work done under the hood of the W3C which may eventually lead to an official specification of various technical aspects of the gadget domain, as well as a number of different gadget APIs. While both are not strictly speaking ontologies, they nevertheless provide a very good insight in how to conceptualise the domain. Additionally, we consider work done in the area of Semantic Web Services, firstly because the gadgets designed in FAST will connect to Semantic and conventional Web services, and secondly because there are a number of similarities between the way SWS are often formalised and the way we envision the interaction of the different components of a gadget in the FAST IDE.

@@

\singlespacing
\begin{small}
\begin{longtable}[t]{|>{\columncolor{fast@lightgrey}}p{3.25cm}|p{11cm}|}
\caption{\label{tab:ontology_requirements_spec}FAST Ontology Requirements Specification Document}\\
\hline
\textcolor{white}{\textbf{Name}} & \textbf{FAST Gadget Ontology} \\ \hline
\textcolor{white}{\textbf{Domain}} & Intelligent Gadgets \\ \hline
\textcolor{white}{\textbf{Purpose}} & The FAST gadget ontology conceptualises the domain of intelligent gadgets as defined in the FAST development platform. Gadgets consist of several interrelated parts, all of which are covered in the ontology. In FAST, a description of each gadget component and resource is available to the IDE. Form this description, the IDE can construct its interface, determine which components can be connected to which other components, or make suggestions to the user in order to aid in the gadget development process.

@@

In summary, the ontology is supposed to facilitate support for the user in the design and generation of FAST gadgets, as well as in searching and browsing those gadgets.\\ \hline 
\textcolor{white}{\textbf{Level of Formality}} & formal (OWL ontology) \\ \hline
\textcolor{white}{\textbf{Scope}} & \textbf{Concepts:} \emph{Component}, \emph{Resource}, \emph{Screen}, \emph{Backend Service}, \emph{Flow Control Element}, \emph{Operator}, \emph{Screenflow Start}, \emph{Screenflow End}, \emph{Connector}, \emph{Form Element}, \emph{Pre-condition}, \emph{Post-condition}, \emph{Query}, \emph{Label}, \emph{Icon}, \emph{User}, \emph{User Profile}, \emph{Tag}

@@

\texttt{Building\_Block} & Anything that is part of a gadget (or the gadget itself). Tentatively anything that can be ``touched'' and moved around in the FAST IDE, from the most complex units such as \texttt{Screen\_Flow}s down to atomic \texttt{Form\_Element}s like a button or a label in a form. \\ \hline
\texttt{Screen\_Flow} & A set of screens from which a gadget for a given target platform can be generated. \\ \hline
\texttt{Screen} & An individual screen; the basic unit of user interaction in FAST. \\ \hline

@@

Regarding their implementation, different kinds of building blocks in FAST can either be defined hard-coded as a piece of source code, or they can be defined declaratively in the terms of the FAST ontology (using pipes, operators, etc.). In the former case, a building block such as a screen would have been implemented and added to the FAST platform by an engineer, whereas in the latter case, ordinary users of the FAST platform would have assembled the building block using the tools available in the GVS\footnote{Currently, only screens can be assembled in the GVS. Forms may or may not be editable in the GVS in the future. Other kinds of building blocks are not planned to be editable.}. This principle is reflected in Fig.~\ref{fig:uml_definition}, showing how certain kinds of building blocks aggregate either a \texttt{Definition} or \texttt{Code}, whereas other kinds of building blocks are always hard-coded.

@@

In this section we will investigate out a number of established ontologies which cover part of the requirements laid out in the specification document, as well as the glossary of terms from the conceptualisation stage. In particular, we will look at the Friend of a Friend (FOAF) ontology, Semantically Interlinked Online Communities (SIOC), the Dublin Core metadata specification and Common Tag. 
The outcome of the integration process is the living \emph{integration document} at \ref{ssub:integration_document}.

\subsubsection{FOAF} % (fold)
\label{ssub:foaf}

@@

Looking at the requirements specification document, DC can be integrated into the FAST gadget ontology in many ways. A lot of the metadata needs for screens, screenflows and their components are identical to those of the media resources originally intended by DC. For this reason, many properties such as \texttt{title}, \texttt{creator}, \texttt{rights} or various \texttt{date} properties can be used directly (more details in the integration document). 
% subsubsection dublin_core (end)

@@

\singlespacing
\begin{small}
\begin{longtable}{|p{3cm}|p{1.5cm}|p{3.5cm}|p{6.1cm}|}
\caption{\label{tab:integration_document}FAST Ontology integration document}\\
\hline
\rowcolor{fast@lightgrey}\textcolor{white}{\textbf{FAST Term}} & 
\textcolor{white}{\textbf{Integrated Ontology}} & 
\textcolor{white}{\textbf{Integrated Term}} & 
\textcolor{white}{\textbf{Comment}} \\ \hline
\endfirsthead
\multicolumn{4}{|c|}{\emph{\textbf{Classes}}} \\ \hline
\texttt{User} & FOAF & \texttt{foaf:Person} & A user in FAST will be modelled using a combination of FOAF and SIOC, so that an instance of \texttt{foaf:Person} will have an account on the FAST platform, represented by an instance of \texttt{sioc:User}, which is a subclass of \texttt{foaf:OnlineAccount}. The relationship is expressed using \texttt{foaf:holdsAccount}. This way of modelling directly follows the suggestions made by~\cite{sioc_related2007}.\\ \hline
\texttt{User} & SIOC & \texttt{sioc:User} & s.a.\\ \hline
\texttt{Image} & FOAF & \texttt{foaf:Image} & --- \\ \hline
\texttt{Tag} & Common Tag & \texttt{ctag:Tag} & Complex or conceptual tags are being used e.g. to model the domain context of resources.	 \\ \hline
\multicolumn{4}{|c|}{\emph{\textbf{Properties}}} \\ \hline
\texttt{hasLabel} & DC & \texttt{dcterms:title} & ---  \\ \hline
\texttt{hasCreator} & DC & \texttt{dcterms:creator} & DC defines the range of \texttt{dcterms:creator} to be \texttt{dcterms:Agent}. By inference, this obviously also holds within FAST. However, we will explicitly only use \texttt{foaf:Person} (also see Fig.~\ref{fig:dcterms_example}). \\ \hline
\texttt{hasDescription} & DC & \texttt{dcterms:description} & --- \\ \hline
\texttt{hasCreationDate} & DC & \texttt{dcterms:created} & Dates should be formatted according to ISO 8601~\cite{w3c_iso8601_1997}. \\ \hline
\texttt{hasRights} & DC & \texttt{dcterms:rights} & DC terms specify a number of classes to model the rights of resources. \texttt{dcterms:rights} links a resource to a \texttt{dcterms:RightsStatement}. The holder of the rights is specified by pointing from the resource to a \texttt{dcterms:Agent}, using the \texttt{dcterms:rightsHolder} property. \\ \hline
\texttt{hasDomainContext} & DC & \texttt{dcterms:subject} & DC specifies that subject shoulds be non-literals. To allow simple tagging, we should represent tags as resources (\todo{common tag}). \\ \hline
\texttt{hasImage} & FOAF & \texttt{foaf:depiction} & The inverse \texttt{foaf:depicts} could also be used. \\ \hline
\texttt{hasIcon} & FOAF & sub-property of \texttt{foaf:depiction} & Icons are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasScreenshot} & FOAF & sub-property of \texttt{foaf:depiction} & Screenshots are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasName} & FOAF & \texttt{foaf:name} & Apart from \texttt{foaf:name}, which is used for the full name, more fine-grained properties such as \texttt{foaf:firstName}, \texttt{foaf:surname} or \texttt{foaf:title} and \texttt{foaf:nick} can be used as well. \\ \hline
\texttt{hasEmail} & FOAF & \texttt{foaf:mbox} & While \texttt{foaf:mbox} will be used for the plain e-mail address, \texttt{foaf:mbox\_sha1sum} will be used for an obfuscated version of the address. \\ \hline
\texttt{hasInterests} & FOAF & \texttt{foaf:interest} & --- \\ \hline
\texttt{hasUserName} & FOAF & \texttt{foaf:accountName} & --- \\ \hline
\texttt{hasDomainContext} & Common Tag & \texttt{ctag:tagged} & The domain context of a resource in FAST can be expressed as a complex tag. Further Common Tag properties can then be applied to make assertions about the tag itself. \\ \hline
\end{longtable}
\end{small}
\doublespacing

@@

@@

This section gives a general, birds-eye-view of the ontology in terms of its classes and properties. Fig.~\ref{fig:ontology_uml} shows all classes in their subsumption hierarchy as a UML diagram, including the properties defining those classes as their domain. A more detailed list of classes and properties, complete with human-readable comments and documentation can be found in the back of the document in App.~\ref{sec:ontology_terms}, while the complete ontology code in OWL is given in App.~\ref{sec:ontology_code}.

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{images/fgo_uml_diagram}
    \caption{FAST Gadget Ontology Class Hierarchy}
    \label{fig:ontology_uml}
  \end{center}
\end{figure}

@@

\begin{itemize}
	\item \textbf{Abstract Class Level}: At this level, resources are defined on a categorical level, differentiating between the various kinds of things that are relevant in FAST, as fundamental classes such as screens, forms or resources (see Sect.~\ref{sub:conceptualisation}). These are implemented technically as classes in OWL.
	\item \textbf{Screenflow-independent Component Definition}: When a user of the FAST GVS is building a new screenflow or screen, they do so by combining existing building blocks and configuring them. These building blocks are not the fundamental classes of FAST, but instead they are specific instances of those classes. In other words, users do not combine the idea of a screen with the idea of a service, but instead they combine the specific ``Log In Screen'' with the ``Amazon Web Service''. Technically, these concrete building blocks are implemented as instances of the fundamental classes, specified by asserting particular settings such as label or icons, pre- and post-conditions, etc.
	\item \textbf{Instantiation within a Screenflow at Design-time}: Once a user selects a specific building block in the GVS and adds it to their new screen or screenflow project, they are in principle again instantiating from a conceptual to a concrete level: from the ``Log In Screen'' as such to a particular log in screen as used in that particular screenflow. Also in this stage, resources are implemented technically as instances.
	\item \textbf{Runtime}: At the final level of instantiation, the screenflow has been packaged as a gadget, deployed to a gadget platform and is now being used by an end-user. At this level, all gadget components exist only in the form of Web standards compliant code (HTML, JavaScript, CSS). Any metadata still necessary or definitions for screen pre- and post-conditions are for performance now defined in the JavaScript native data format JSON, rather than in RDF. Therefore, at this level, the resources that make up the gadget are out of the scope of this deliverable.
\end{itemize}

@@

When implementing these four stages, as illustrated in Fig.~\ref{fig:classes_templates}, to OWL DL semantics, one will be faced with an immediate problem. Conceptually, once would like to say that $Screen$ is a class, i.e., the set of all screens. As argued above, $LogInScreen$ would then be an instance of $Screen$ ($LogInScreen \subset Screen$), while a particular $LogInScreen_x$ in a particular screenflow would again be an instance of $LogInScreen$ ($LogInScreen_x \subset LogInScreen$). However, unless we want to move into OWL Full semantics, classes cannot be used as instances at the same time, so we cannot use this modelling approach. Furthermore, when we instantiate a resource such as a screen within a particular screenflow, we want to imply that the new instance ($LogInScreen_x$) has the same features such as label, icon or conditions, as $LogInScreen$. In other words, what we really require is a something like a \emph{copy} of the original $LogInScreen$.

Neither OWL nor RDFS semantics offer functionality to represent a relationship such as the copy of a resource formally. For this reason, we introduce the concept of \emph{templates for RDF instances} in FAST. One resource (e.g., $LogInScreen$) will play the role of a \emph{template}, while the other resource (e.g., $LogInScreen_x$) will be called a \emph{copy}. Conceptually, copying a resource means defining a named graph that contains all statements about this resource considered to be relevant, and creating a new graph with near-identical statements\footnote{The graphs are not completely identical: the identifiers for the resource and potentially other, contained resources will have changed.}. In practice, this procedure can be implemented using a SPARQL CONSTRUCT query, where the WHERE clause matches the graph pattern of the template, and the CONSTRUCT query will build the copy from it. The corresponding data on all three levels is shown in List.~\ref{list:templating}, using TriG syntax~\cite{bizer2004trig} (an extension of Turtle syntax for named graphs).

@@

In this section, we will present the use of some basic resource annotation properties in FAST. In general, since screenflows, screens and their components are sub-classes of \htmlref{\texttt{fgo:Resource}}{subs:Resource}, all properties defined to have a domain of \htmlref{\texttt{fgo:Resource}}{subs:Resource} should be used. In addition, a number of terms from external ontologies should also be used to annotate resources. Note that all of these annotations are valid both for the screenflow-independent component definitions (templates) and the design time instances (copies). A detailed description of all terms from the FAST gadget ontology namespace can be found in Sect.~\ref{sec:ontology_terms} of this document.

@@

@@

Since graph patterns with variable cannot directly be expressed in RDF (short of using blank nodes, which cannot have labels), the graph patterns are expressed as RDF literals and linked \texttt{fast:Condition} instances using the \texttt{fast:hasPattern} property.
Each such condition will be linked to a screen or screenflow using either the \texttt{fast:hasPreCondition} or \texttt{fast:hasPostCondition} property.

@@

\begin{itemize}
	\item \emph{Domain Contexts:} By providing a domain context for a screen (or any other FAST resource), it is possible to express what domain this resource is relevant for. Examples for domain contexts are ``medicine'', ``EU projects'', ``catering'', ``human resources'', ``books'', etc. As defined in the Integration Document in Sect.~\ref{ssub:integration_document}, domain contexts are given using the \emph{dcterms:subject} property. One option for finding domain identifiers to be used with this property is the use of DBpedia~\cite{Auer07dbpedia} identifiers such as \url{http://dbpedia.org/resource/Medicine} (DBpedia is a large RDF corpus automatically extracted from the info boxes of Wikipedia entries). This is now a widely used and recommended practice in the Semantic Web community (similar corpora could also be used). As an alternative or supplement to DBpedia identifiers, it is possible to use structured tagging approaches such as MOAT\footnote{\url{http://moat-project.org/}}.
	\todo{update with reference to common tag!}
	\item \emph{User Interests:} In the same way as domain contexts, a user's interests can be specified using DBpedia identifiers or structured tagging. Both can then be used in combination to suggest resources to the user.
	\item \emph{Pre- and Post-conditions:} In specifying pre- and post-conditions, terminology for an open-ended number of domains is necessary. While a number of classes will be reused quite often, very specific screens will require very specific vocabulary. We cannot, as part of the FAST project, provide vocabulary for all conceivable domains of discourse, and while also here DBpedia identifiers may be an option, often more precise terms will be needed.
\end{itemize}

It should be added that for any given screenflow, the concrete definition of both domain context and could be derived automatically from any semantic description which the backend services might have. While this might happen directly, without any transformation or mapping, this approach would be impractical in practice. The reason for this is that each backend service can potentially (and very probably) express its semantics in different ontologies or vocabularies, which would render the pre- and post-condition mechanism in FAST useless. Therefore, we assume that a mapping or mediation layer will ensure that the different semantics at the backend service level will be mapped to a common set of terms within the FAST platform. This problem is the topic of deliverable 2.4 in FAST~\cite{ambrus2010fast_mediation} and has also been discussed in \cite{Ambrus:2009it}.
% subsection extension_towards_specific_domains (end)

@@

In the following, we will look at each of those guidelines and discuss in how far the FAST gadget ontology fulfils them.

@@

The concrete interpretation of this criterion depends to a large degree on the intended use case and user community of the ontology. For a large-scale undertaking spanning many domains, and where reasoning is in the centre of attention --- CyC e.g. represents this type of ontology  --- a certain amount of complexity is necessary. However, in order to appeal to adopters who are familiar with formal knowledge representation, simplicity can be an important factor. Good examples of such ontologies are FOAF and SIOC, which would arguably not have had such an impact in the wider Web community if they tried to model their respective domains in a much more complex way, making use of the full range of expressivity offered by e.g. the RDFS and OWL ontology languages. Indirect support for this argument is provided by \cite{wang2006owl_survey}, who show that most ontologies found on the Web are on the lower end of the expressivity spectrum, indicating that practitioners either don't need or don't understand the full range of possible expressiveness.

>>>>

% This helps to get hyphenation in typewriter font (\texttt)
\usepackage[htt]{hyphenat}

@@

This deliverable is structured as follows: as part of the introduction, we will present the methodology we chose to adopt for the development of the FAST gadget ontology, by discussing  each of its proposed stages in detail. Following, we will conclude the introduction by elaborating more on some of the general design decisions we are following. After briefly addressing the topic of related work in Sect.~\ref{sec:related_work} --- which is mainly covered in a different deliverable from this work package, \cite{urmetzer2010fast_state_of_the_art} --- the structure of the remainder of this document follows directly from our adopted methodology: we will perform a domain analysis in Sect.~\ref{sec:domain_analysis}, covering the stages of specification, conceptualisation and integration of existing vocabularies. This is then followed by the implementation step in Sect.~\ref{sec:ontology}, which covers the ontology in its form at this moment of the project. Finishing the deliverable, we present our conclusions and outlook on years two and three of the project in Sect.~\ref{sec:conclusions}.

@@

\todo{update this section, refer to change requests in last review}

@@

We are not aware of any other project which has already developed or planned to develop an ontology of the gadget domain. However, there is a range of related material from which we take inspiration regarding the description of the gadget domain. This material comes from a number of different sources and directions. It includes work done under the hood of the W3C which may eventually lead to an official specification of various technical aspects of the gadget domain, as well as a number of different gadget APIs. While both are not strictly speaking ontologies, they nevertheless provide a very good insight in how to conceptualise the domain. Additionally, we consider work done in the area of Semantic Web Services, firstly because the gadgets designed in FAST will eventually connect to Semantic and conventional Web services, and secondly because there are a number of similarities between the way SWS are often formalised and the way we envision the interaction of the different components of a gadget in the FAST IDE.

@@

\singlespacing
\begin{small}
\begin{longtable}[t]{|>{\columncolor{fast@lightgrey}}p{3.25cm}|p{11cm}|}
\caption{\label{tab:ontology_requirements_spec}FAST Ontology Requirements Specification Document}\\
\hline
\textcolor{white}{\textbf{Name}} & \textbf{FAST Gadget Ontology} \\ \hline
\textcolor{white}{\textbf{Domain}} & Intelligent and complex gadgets \\ \hline
\textcolor{white}{\textbf{Purpose}} & The FAST gadget ontology conceptualises the domain of intelligent gadgets as defined in the FAST development platform. Gadgets consist of several interrelated parts, all of which are covered in the ontology. In FAST, a description of each gadget component and resource is available to the IDE. From this description, the IDE can construct its interface, determine which components can be connected to which other components, or make suggestions to the user in order to aid in the gadget development process.

@@

In summary, the ontology is supposed to facilitate support for the user in the design and generation of FAST gadgets, as well as in searching and browsing those gadgets.\\ \hline 
\textcolor{white}{\textbf{Level of Formality}} & formal (OWL ontology) \\ \hline
\textcolor{white}{\textbf{Scope}} & \textbf{Concepts:} \emph{Component}, \emph{Resource}, \emph{Screen}, \emph{Screen flow}, \emph{Backend Service}, \emph{Flow Control Element}, \emph{Operator}, \emph{Screenflow Start}, \emph{Screenflow End}, \emph{Connector}, \emph{Form Element}, \emph{Pre-condition}, \emph{Post-condition}, \emph{Query}, \emph{Label}, \emph{Icon}, \emph{User}, \emph{User Profile}, \emph{Tag}

@@

\texttt{Building\_Block} & Anything that is part of a gadget. Tentatively anything that can be ``touched'' and moved around in the FAST IDE, from the most complex units such as \texttt{Screen\_Flow}s, down to atomic \texttt{Form\_Element}s like a button or a label in a form. \\ \hline
\texttt{Screen\_Flow} & A set of screens from which a gadget in a given format (which can cover one or more target platforms) can be generated. \\ \hline
\texttt{Screen} & An individual screen; the basic unit of user interaction in FAST. A screen is the interface through which a user gets access to data and functionality of a backend service. \\ \hline

@@

Regarding their implementation, different kinds of building blocks in FAST can either be defined hard-coded as a piece of source code, or they can be defined declaratively in the terms of the FAST ontology (using pipes, operators, etc.). In the former case, a building block such as a screen would have been implemented and added to the FAST platform by an engineer, whereas in the latter case, ordinary users of the FAST platform would have assembled the building block using the tools available in the GVS\footnote{At this moment, only screens can be assembled in the GVS. Forms may or may not be editable in the GVS in the future. Other kinds of building blocks are not currently planned to be editable.}. This principle is reflected in Fig.~\ref{fig:uml_definition}, showing how certain kinds of building blocks aggregate either a \texttt{Definition} or \texttt{Code}, whereas other kinds of building blocks are always hard-coded. It should be noted that, in theory, there is no reason why not any kind of building block could be defined either way. However, the concrete set of building blocks which can be defined declaratively or in code is changing dynamically according to the current state and plans of the FAST project. The situation depicted in Fig.~\ref{fig:uml_definition} --- \texttt{Screen} and any kind of \texttt{Screen\_Component} can be defined in code, whereas only \texttt{Screen} and \texttt{Form} can be defined declaratively --- reflects the status at M24 of the project.

@@

In this section we will investigate a number of established ontologies which cover part of the requirements laid out in the specification document, as well as the glossary of terms from the conceptualisation stage. In particular, we will look at the Dublin Core metadata specification, the Friend of a Friend (FOAF) ontology, Semantically Interlinked Online Communities (SIOC)  and Common Tag. 
The outcome of the integration process is the living \emph{integration document} at \ref{ssub:integration_document}.

Looking at the requirements specification document, DC can be integrated into the FAST gadget ontology in many ways. A lot of the metadata needs for screens, screenflows and their components are identical to those of the media resources originally intended by DC. For this reason, many properties such as \texttt{title}, \texttt{creator}, \texttt{rights} or various \texttt{date} properties can be used directly (more details in the integration document). 
% subsubsection dublin_core (end)
\subsubsection{FOAF} % (fold)
\label{ssub:foaf}

@@

@@

\singlespacing
\begin{small}
\begin{longtable}{|p{3cm}|p{1.5cm}|p{3.5cm}|p{6.1cm}|}
\caption{\label{tab:integration_document}FAST Ontology integration document}\\
\hline
\rowcolor{fast@lightgrey}\textcolor{white}{\textbf{FAST Term}} & 
\textcolor{white}{\textbf{Integrated Ontology}} & 
\textcolor{white}{\textbf{Integrated Term}} & 
\textcolor{white}{\textbf{Comment}} \\ \hline
\endfirsthead
\multicolumn{4}{|c|}{\emph{\textbf{Classes}}} \\ \hline
\texttt{User} & FOAF & \texttt{foaf:Person} & A user in FAST will be modelled using a combination of FOAF and SIOC, so that an instance of \texttt{foaf:Person} will have an account on the FAST platform, represented by an instance of \texttt{sioc:User}, which is a subclass of \texttt{foaf:OnlineAccount}. The relationship is expressed using \texttt{foaf:holdsAccount}. This way of modelling directly follows the suggestions made by~\cite{sioc_related2007}.\\ \hline
\texttt{User} & SIOC & \texttt{sioc:User} & s.a.\\ \hline
\texttt{Image} & FOAF & \texttt{foaf:Image} & --- \\ \hline
\texttt{Tag} & Common Tag & \texttt{ctag:Tag} & Complex or conceptual tags are being used e.g. to model the domain context of resources.	 \\ \hline
\multicolumn{4}{|c|}{\emph{\textbf{Properties}}} \\ \hline
\texttt{hasLabel} & DC & \texttt{dcterms:title} & ---  \\ \hline
\texttt{hasCreator} & DC & \texttt{dcterms:creator} & DC defines the range of \texttt{dcterms:creator} to be \texttt{dcterms:Agent}. By inference, this obviously also holds within FAST. However, we will explicitly only use \texttt{foaf:Person} (also see Fig.~\ref{fig:dcterms_example}). \\ \hline
\texttt{hasDescription} & DC & \texttt{dcterms:description} & --- \\ \hline
\texttt{hasCreationDate} & DC & \texttt{dcterms:created} & Dates should be formatted according to ISO 8601~\cite{w3c_iso8601_1997}. \\ \hline
\texttt{hasRights} & DC & \texttt{dcterms:rights} & DC terms specify a number of classes to model the rights of resources. \texttt{dcterms:rights} links a resource to a \texttt{dcterms:RightsStatement}. The holder of the rights is specified by pointing from the resource to a \texttt{dcterms:Agent}, using the \texttt{dcterms:rightsHolder} property. \\ \hline
% \texttt{hasDomainContext} & DC & \texttt{dcterms:subject} & DC specifies that subject shoulds be non-literals. To allow simple tagging, we should represent tags as resources (\todo{common tag}). \\ \hline
\texttt{hasImage} & FOAF & \texttt{foaf:depiction} & The inverse \texttt{foaf:depicts} could also be used. \\ \hline
\texttt{hasIcon} & FOAF & sub-property of \texttt{foaf:depiction} & Icons are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasScreenshot} & FOAF & sub-property of \texttt{foaf:depiction} & Screenshots are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasName} & FOAF & \texttt{foaf:name} & Apart from \texttt{foaf:name}, which is used for the full name, more fine-grained properties such as \texttt{foaf:firstName}, \texttt{foaf:surname} or \texttt{foaf:title} and \texttt{foaf:nick} can be used as well. \\ \hline
\texttt{hasEmail} & FOAF & \texttt{foaf:mbox} & While \texttt{foaf:mbox} will be used for the plain e-mail address, \texttt{foaf:mbox\_sha1sum} will be used for an obfuscated version of the address. \\ \hline
\texttt{hasInterests} & FOAF & \texttt{foaf:interest} & --- \\ \hline
\texttt{hasUserName} & FOAF & \texttt{foaf:accountName} & --- \\ \hline
\texttt{hasDomainContext} & Common Tag & \texttt{ctag:tagged} & The domain context of a resource in FAST can be expressed as a complex tag. Further Common Tag properties can then be applied to make assertions about the tag itself. \\ \hline
\texttt{hasVersion} & DOAP & \texttt{daop:revision} & DOAP (Description of a Project) is a widely used vocabulary for basic annotation of software and other projects. \\ \hline
\end{longtable}
\end{small}
\doublespacing

@@

\todo{add graph figures to all examples}

@@

This section gives a general, birds-eye-view of the ontology in terms of its classes and properties. Fig.~\ref{fig:ontology_uml} shows all classes in their subsumption hierarchy as a UML diagram, including the properties defining those classes as their domain. All \texttt{rdfs:subClassOf}/\texttt{rdfs:superClassOf} relations are modelled with the subsumption arrow, while \texttt{owl:union\_of} relations are modelled using the filled composition diamond.
A more detailed list of classes and properties, complete with human-readable comments and documentation can be found in the back of the document in App.~\ref{sec:ontology_terms}, while the complete ontology code in OWL is given in App.~\ref{sec:ontology_code}.

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{images/fgo_uml_diagram}
    \caption{FAST Gadget Class Hierarchy}
    \label{fig:ontology_uml}
  \end{center}
\end{figure}

@@

\begin{itemize}
	\item \textbf{Abstract Class Level}: At this level, resources are defined on a categorical level, differentiating between the various kinds of things that are relevant in FAST, as fundamental classes such as screens, forms or resources (see Sect.~\ref{sub:conceptualisation}). These are implemented technically as classes in OWL.
	\item \textbf{Prototypical Component Definition}: When a user of the FAST GVS is building a new screenflow or screen, they do so by combining existing building blocks and configuring them. These building blocks are not the fundamental classes of FAST, but instead they are specific instances of those classes. In other words, users do not combine the idea of a screen with the idea of a service, but instead they combine the specific ``Log In Screen'' with the ``Amazon Web Service''. Technically, these concrete building blocks are implemented as instances of the fundamental classes, specified by asserting particular settings such as label or icons, pre- and post-conditions, etc.
	\item \textbf{Instantiation within a Screenflow at Design-time}: Once a user selects a specific building block in the GVS and adds it to their new screen or screenflow project, they are in principle again instantiating from a conceptual to a concrete level: from the ``Log In Screen'' as such to a particular log in screen as used in that particular screenflow. Also in this stage, resources are implemented technically as instances.
	\item \textbf{Runtime}: At the final level of instantiation, the screenflow has been packaged as a gadget, deployed to a gadget platform and is now being used by an end-user. At this level, all gadget components exist only in the form of Web standards compliant code (HTML, JavaScript, CSS). Any metadata still necessary or definitions for screen pre- and post-conditions are for performance now defined in the JavaScript-native data format JSON, rather than in RDF. Therefore, at this level, the resources that make up the gadget are out of the scope of this deliverable.
\end{itemize}

@@

When implementing these four stages, as illustrated in Fig.~\ref{fig:classes_templates}, to OWL DL semantics, one will be faced with an immediate problem. Conceptually, one would like to say that $Screen$ is a class, i.e., the set of all screens. As argued above, $LogInScreen$ would then be an instance of $Screen$ ($LogInScreen \subset Screen$), while a particular $LogInScreen_x$ in a particular screenflow would again be an instance of $LogInScreen$ ($LogInScreen_x \subset LogInScreen$). However, unless we want to move into OWL Full semantics, classes cannot be used as instances at the same time, so we cannot use this modelling approach. Furthermore, when we instantiate a resource such as a screen within a particular screenflow, we want to imply that the new instance ($LogInScreen_x$) has the same features such as label, icon or conditions, as $LogInScreen$. In other words, what we really require is a something like a \emph{copy} of the original $LogInScreen$.

Neither OWL nor RDFS semantics offer functionality to represent a relationship such as the copy of a resource formally. For this reason, we introduce the concept of \emph{templates for RDF instances} in FAST. One resource (e.g., $LogInScreen$) will play the role of a \emph{template}, while the other resource (e.g., $LogInScreen_x$) will be called a \emph{copy}. Conceptually, copying a resource means defining a named graph that contains all statements about this resource considered to be relevant, and creating a new graph with near-identical statements\footnote{The graphs are not completely identical: the identifiers for the resource and potentially other, contained resources will have changed.} \todo{explain this better}. In practice, this procedure can be implemented using a SPARQL CONSTRUCT query, where the WHERE clause matches the graph pattern of the template, and the CONSTRUCT query will build the copy from it. The corresponding data on all three levels is shown in List.~\ref{list:templating}, using TriG syntax~\cite{bizer2004trig} (an extension of Turtle syntax for named graphs).

@@

In this section, we will present the use of some basic resource annotation properties in FAST. In general, since screenflows, screens and their components are sub-classes of \htmlref{\texttt{fgo:Resource}}{subs:Resource}, all properties defined to have a domain of \htmlref{\texttt{fgo:Resource}}{subs:Resource} should be used. In addition, a number of terms from external ontologies should also be used to annotate resources. Note that all of these annotations are valid both for the prototypical component definitions (templates) and the design time instances (copies). A detailed description of all terms from the FAST gadget ontology namespace can be found in Sect.~\ref{sec:ontology_terms} of this document.

@@

\todo{change examples to good relations}

@@

Since graph patterns with variable cannot directly be expressed in RDF (short of using blank nodes, which cannot have labels), the graph patterns are expressed as RDF literals and linked \texttt{fast:Condition} instances using the \texttt{fast:hasPattern} property.
Each such condition will be linked to a screen or screenflow using either the \texttt{fast:has\-PreCondition} or \texttt{fast:has\-Post\-Condition} property.

@@

\begin{itemize}
	\item \emph{Domain Contexts:} By providing a domain context for a screen (or any other FAST resource), it is possible to express what domain this resource is relevant for. Examples for domain contexts are ``medicine'', ``EU projects'', ``catering'', ``human resources'', ``books'', etc. As defined in the Integration Document in Sect.~\ref{ssub:integration_document}, domain contexts are given using the \emph{dcterms:subject} property. One option for finding domain identifiers to be used with this property is the use of DBpedia~\cite{Auer07dbpedia} identifiers such as \url{http://dbpedia.org/resource/Medicine} (DBpedia is a large RDF corpus automatically extracted from the info boxes of Wikipedia entries). This is now a widely used and recommended practice in the Semantic Web community (similar corpora could also be used). As an alternative or supplement to DBpedia identifiers, it is possible to use structured tagging approaches such as MOAT\footnote{\url{http://moat-project.org/}}.
	\todo{update with reference to common tag!}
	\todo{give concrete example with code, maybe Good Relations?}
	\item \emph{User Interests:} In the same way as domain contexts, a user's interests can be specified using DBpedia identifiers or structured tagging. Both can then be used in combination to suggest resources to the user.
	\item \emph{Pre- and Post-conditions:} In specifying pre- and post-conditions, terminology for an open-ended number of domains is necessary. While a number of classes will be reused quite often, very specific screens will require very specific vocabulary. We cannot, as part of the FAST project, provide vocabulary for all conceivable domains of discourse, and while also here DBpedia identifiers may be an option, often more precise terms will be needed.
\end{itemize}

It should be added that for any given screenflow, the concrete definition of both domain context could be derived automatically from any semantic description which the backend services might have. While this might happen directly, without any transformation or mapping, this approach would be impractical in practice. The reason for this is that each backend service can potentially (and very probably) express its semantics in different ontologies or vocabularies, which would render the pre- and post-condition mechanism in FAST useless. Therefore, we assume that a mapping or mediation layer will ensure that the different semantics at the backend service level will be mapped to a common set of terms within the FAST platform. This problem is the topic of deliverable 2.4 in FAST~\cite{ambrus2010fast_mediation} and has also been discussed in \cite{Ambrus:2009it}.
% subsection extension_towards_specific_domains (end)

@@

In the following, we will look at each of those guidelines and discuss how far the FAST gadget ontology fulfils them.

@@

The concrete interpretation of this criterion depends to a large degree on the intended use case and user community of the ontology. For a large-scale undertaking spanning many domains, and where reasoning is in the centre of attention --- CyC\footnote{\url{http://cyc.com/cyc/opencyc}} e.g. represents this type of ontology  --- a certain amount of complexity is necessary. However, in order to appeal to adopters who are familiar with formal knowledge representation, simplicity can be an important factor. Good examples of such ontologies are FOAF and SIOC, which would arguably not have had such an impact in the wider Web community if they tried to model their respective domains in a much more complex way, making use of the full range of expressivity offered by e.g. the RDFS and OWL ontology languages. Indirect support for this argument is provided by \cite{wang2006owl_survey}, who show that most ontologies found on the Web are on the lower end of the expressivity spectrum, indicating that practitioners either don't need or don't understand the full range of possible expressiveness.
